---
title: "Geospatial Mapping for Ornithology"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## 1) Load packages we'll use

```{r}
library(tidyverse)
library(ggmap)
library(sf)
library(elevatr)
library(raster)
library(marmap)
```

## 2) Build a basemap

### a) ...using a satellite image

#### Register Google API key

To obtain an API key and enable services, go to <https://mapsplatform.google.com>. (Also note that your API key should be kept secure, as sharing it publicly can open your account up to being compromised.)

My code chunk is hidden here, but the function to register your API key in an R session takes the form `register_google(key = "PERSONAL API KEY")`.

```{r, include=FALSE}
register_google(key = "AIzaSyBn1OxFq4DMK79J4qHmQoBvFlbHbeImABw")
```

#### Extract satellite map from Google Maps

Right click on a Google Maps page to get coordinates. Also note that _longitude_ comes first inside the coordinate parentheses.

```{r}
island_map <- get_map(location = c(-70.614, 42.9885),
                      zoom = 16,
                      maptype = "satellite",
                      source = "google",
                      color = "color") # use "bw" for a black-and-white satellite image
```

#### Create basemap

When transforming the earth from its 3-D spherical shape to a 2-D shape, various distinct projections may be used depending on the scale and purposes of a given map. A Coordinate Reference System (CRS) defines how a specific two-dimensional projection relates to real places on the earth, and therefore is an important specification in geospatial mapping. Here, we use the most popular geodetic datum, WGS84 (World Geodetic System 1984), represented by the EPSG code 4326.

Within `get_map()`, feel free to experiment with the other `maptype` options for maps sourced from Google or Stamen, examples here: <https://r-graph-gallery.com/324-map-background-with-the-ggmap-library.html>.

```{r}
satmap <- ggmap(island_map) +
  coord_sf(crs = 4326) +
  theme(axis.title = element_blank())

# note that the CRS is set in `coord_sf()` solely for the sake of formatting the axis labels; the satellite map is already set to WGS84/EPSG:4326.

satmap
```

#### Zoom in a bit more by setting coordinate axis limits

```{r}
satmap <- satmap +
  xlim(c(-70.619, -70.6094)) +
  ylim(c(42.9842, 42.9928))

satmap
```

### b) ...using topographic data

The `elevatr` package provides access to raster elevation data from AWS Open Data [Terrain Tiles](https://registry.opendata.aws/terrain-tiles/) and the Open Topography [Global datasets API](https://opentopography.org/developers#API).

#### Prepare `locations` parameter

This is the object that we will use to crop the global elevation data.

"Simple Features" is a commonly used set of GIS standards for storing and accessing data representing predominantly two-dimensional geographic features (e.g. points, lines, polygons).

```{r}
# choose map GPS limits
min <- c(-70.619, 42.9842)
max <- c(-70.6094, 42.9928)

# create dataframe for bounding box
loc <- rbind(min, max) %>%
  as.data.frame() %>%
  rename(Long = V1, Lat = V2)

# convert dataframe into simple features object for mapping
  # this serves to assign the data a CRS
loc_sf <- st_as_sf(loc,
                   coords = c("Long", "Lat"),
                   crs = 4326)
```

#### Fetch elevation data

```{r}
elevation <- get_elev_raster(locations = loc_sf,
                             z = 14,
                             clip = "bbox")

# convert raster data to point data for plotting
elevation <- rasterToPoints(elevation)

elevation_df <- as.data.frame(elevation) %>%
  na.omit()

# rename the third column to make it easier to reference
colnames(elevation_df)[3] <- "elevation"
```

#### Plot topographic map

```{r}
ggplot() +
  geom_raster(data = elevation_df, 
  			aes(x = x, y = y, fill = elevation))
```

#### Change colour scale

```{r}
# set all sea level points to the same value to create uniform background
for(i in 1:nrow(elevation_df)) {
  if(elevation_df$elevation[i] <= 0.1) {
    elevation_df$elevation[i] = 0
  }
}

elevmap <- ggplot() +
  geom_raster(data = elevation_df, 
  			aes(x = x, y = y, fill = elevation)) +
  # change colour scale
    # "rev" reverses the colour palette to make blue the color for low elevations
    # (100) sets the number of individual values in the scale to create a smooth gradient
    # [30:89] chooses a subset of colors from the palette
  scale_fill_gradientn(colours = rev(etopo.colors(100)[30:89]))

elevmap
```

#### Set CRS to fix axis proportions; add other edits

```{r}
elevmap <- elevmap +
  coord_sf(crs = 4326, expand = FALSE) +
  theme(axis.title = element_blank(),
        legend.position = "none")

elevmap
```

## 3) Import and plot point data

#### Import nest data

```{r}
nests <- read.csv("~/Desktop/Repos/wos2023-r-geospatial/workshop-files/2022_GBBGnests.csv")
```

#### Optional: create a simple features object from nest data

```{r}
nests_sf <- st_as_sf(nests,
                     coords = c("Long", "Lat"),
                     crs = 4326)
```

#### Plot points on a satellite map

```{r}
satmap +
  geom_point(data = nests, aes(x = Long, y = Lat))
  # use the line below if you are adding points with simple features geometry
  #geom_sf(data = nests_sf, inherit.aes = FALSE)
```

#### Change point colour and style
```{r}
satmap <- satmap +
  geom_point(data = nests, aes(x = Long, y = Lat), shape = 21, fill = "white", size = 2.5)
  #geom_sf(data = nests_sf, inherit.aes = FALSE)

satmap
```

#### Zoom into a subsection of the map

```{r}
satmap +
  coord_sf(xlim = c(-70.617, -70.6105), ylim = c(42.988, 42.990), crs = 4326)
```

#### Vary point colour with another variable

```{r}
satmap +
  geom_point(data = nests, aes(x = Long, y = Lat, fill = factor(Eggs)), shape = 21, size = 2.5) +
  scale_fill_manual(values = c("dodgerblue", "goldenrod1", "white"), name = "Number\nof Eggs") +
  theme(legend.title = element_text(face = "bold"))
```

#### Plot points on a topographic map

```{r}
elevmap +
  geom_point(data = nests, aes(x = Long, y = Lat), shape = 21, fill = "white", size = 2.5)
```

## 4) Other possibilities

For these final examples, we will use a dataset of fishing sites on Georges Bank in the north Atlantic.

#### First, create the basemap we'll use

```{r}
# import bathymetric data from NOAA's ETOPO 2022 database
GB_bathy <- getNOAA.bathy(lon1 = -74,
                          lon2 = -62,
                          lat1 = 37,
                          lat2 = 46,
                          resolution = 1)

# prepare bathy data for mapping with ggplot
ggbathy <- GB_bathy %>%
  fortify() %>%
  # create custom depth contours
  mutate(depth_bins = cut(z,
                          breaks = c(0, -30, -55, -75, -90, -120, -150, -180,
                                     -780, -1380, -1980, -2580, -3180, -Inf)))

Georges <- ggplot() +
  geom_raster(data = ggbathy,
              aes(x, y, fill = depth_bins),
              interpolate = TRUE,
              alpha = 0.75) +
  coord_sf(xlim = c(-70.5, -66), ylim = c(40, 42.5), crs = st_crs(4326)) +
  theme(axis.title = element_blank(),
        legend.position = "none")

Georges

# change default colour scale to shades of blue
Georges <- Georges +
  scale_fill_manual(values = c("#08306B", "#084184", "#08519C", "#1561A9",
                               "#2171B5", "#3282BE", "#4292C6", "#57A0CE",
                               "#6BAED6", "#85BCDC", "#9ECAE1", "#B2D3E8",
                               "#C6DBEF"))

Georges
```

### a) Add high-resolution coastline

GSHHG (Global Self-consistent, Hierarchical, High-resolution Geography) shoreline data can be downloaded from <https://www.ngdc.noaa.gov/mgg/shorelines/data/gshhg/latest>.

Spherical geometry is turned off because many polygons cannot be valid on S2 and R2 simultaneously, as outlined in <https://r-spatial.org/book/04-Spherical.html#validity-on-the-sphere>. (This is an excellent book for general referencing!)

The `st_make_valid()` function serves to resolve any areas of geometry that are not topologically valid and thus would raise a GEOS error. Another useful function not needed here, `st_remove_holes()` (from the package `nngeo`), similarly resolves issues such as false slivers of space between contiguous polygons.

```{r}
# turn off spherical geometry to avoid "duplicate vertex" errors
sf_use_s2(FALSE)

# read shapefile from wherever the file is located
mass <- read_sf("/Library/Frameworks/R.framework/Versions/4.2-arm64/Resources/library/gshhg-shp-2.3.7/GSHHS_shp/f/GSHHS_f_L1.shp") %>%
  st_make_valid() %>%
  st_crop(xmin = -71, xmax = -66, ymin = 40, ymax = 43)

# add to map
Georges <- Georges +
  geom_sf(data = mass, fill = "darkgrey", colour = "black") +
  scale_fill_manual(values = c("#08306B", "#084184", "#08519C", "#1561A9",
                               "#2171B5", "#3282BE", "#4292C6", "#57A0CE",
                               "#6BAED6", "#85BCDC", "#9ECAE1", "#B2D3E8",
                               "#C6DBEF"),
                    na.value = "#C6DBEF") +   # cleans up pixelated coastline edges
  coord_sf(xlim = c(-70.5, -66), ylim = c(40, 42.5)) +
  scale_y_continuous(breaks = c(40, 41, 42))

Georges
```

### b) Import and plot spatial polygons

#### Import scallop fishing zone data

```{r}
GB <- read_sf("~/Desktop/Repos/wos2023-r-geospatial/workshop-files/2020SAMZones/GB_Estimation_Areas_2019_UTM19_PDT_SFModified.shp")

# subset open fishing areas
GSC <- subset(GB, NewSAMS == "GSC")
NF <- subset(GB, NewSAMS == "NF")
CL2_AccessSoutheast <- subset(GB, NewSAMS == "CL2-Access-Southeast")
NLS_West <- subset(GB, NewSAMS == "NLS-West")
```

Note that the innate CRS of this polygon collection is NAD83 / UTM zone 19N (which is EPSG:26919), *not* WGS 84; therefore it is important that we set the CRS to EPSG:4326 in `coord_sf()` in order to align with our existing basemap.

#### Plot polygons

```{r}
Georges +
  geom_sf(data = GSC, fill = NA, colour = "black",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = NF, fill = NA, colour = "black",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = CL2_AccessSoutheast, fill = NA, colour = "black",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = NLS_West, fill = NA, colour = "black",
          linewidth = 0.4, linetype = "dashed") +
  coord_sf(xlim = c(-70.5, -66), ylim = c(40, 42.5), crs = st_crs(4326))
```

#### Colour polygons

```{r}
Georges <- Georges +
  geom_sf(data = GSC, fill = "firebrick1", alpha = 0.5, colour = "black",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = NF, fill = "firebrick1", alpha = 0.5, colour = "black",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = CL2_AccessSoutheast, fill = "firebrick1", alpha = 0.5, colour = "black",
          linewidth = 0.4, linetype = "dashed") +
  geom_sf(data = NLS_West, fill = "firebrick1", alpha = 0.5, colour = "black",
          linewidth = 0.4, linetype = "dashed") +
  coord_sf(xlim = c(-70.5, -66), ylim = c(40, 42.5), crs = st_crs(4326))

Georges
```

#### Import and plot point data

```{r}
fishing <- read.csv("~/Desktop/Repos/wos2023-r-geospatial/workshop-files/OctFishingData.csv") %>%
  rename(Lat = Latitude_degrees_start,
         Long = Longitude_degrees_start)

Georges +
  geom_point(data = fishing, aes(x = Long, y = Lat), shape = 17, size = 2)
```

### c) Vary point size with another variable

```{r}
Georges +
  geom_point(data = fishing, aes(x = Long, y = Lat, size = Tow_abundance), shape = 17)
```

## The end!

Thanks so much for viewing! Please feel free to reach out to me at `dylan.titmuss@whoi.edu` with any questions or other thoughts. :)
